BIP-Predictor

### 2025-05-21 Sweave/BIP_predictor.Rnw
up

# Run the M-SSA wrapper, see tutorial 7.2
#   -The function computes M-SSA and M-MSE for each forecast horizon h in h_vec
if (recompute_results)
{
  final_mssa_indicator_obj<-compute_mssa_BIP_predictors_func(x_mat_wc,lambda_HP,L,date_to_fit_all,p,q,ht_mssa_vec,h_vec,f_excess,lag_vec,select_vec_multi)
  save(final_mssa_indicator_obj,file=paste(path.result,"/final_mssa_indicator_obj",sep=""))
} else
{
  load(file=paste(path.result,"/final_mssa_indicator_obj",sep=""))
}
# Final M-SSA components
final_mssa_array<-final_mssa_indicator_obj$mssa_array

@@

colo<-c("blue","red")
mplot<-scale(mssa_final_vs_prem_wc)
mplot<-mplot[L:nrow(mplot),]
colnames(mplot)<-colnames(mssa_final_vs_prem_wc)
main_title<-paste("GDP component predictors: final and 2008-MSSA (h=",h,", shift=",shift,")",sep="")
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
#mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (jj in 1:ncol(mplot))
{
  lines(mplot[,jj],col=colo[jj],lwd=1,lty=1)
#  mtext(colnames(mplot)[jj],col=colo[jj],line=-jj)
}
abline(h=0)
axis(1,at=c(1,12*1:(nrow(mplot)/12)),labels=rownames(mplot)[c(1,12*1:(nrow(mplot)/12))])
axis(2)
legend("bottomleft", legend = colnames(mplot), col = colo, lty = 1, lwd = 1, bty = "n")

@@

@@

>>>>

# We can select type of VAR model
VAR_type="BVAR"
VAR_type="VAR"
if (VAR_type=="BVAR")
{
# Simon's settings for BVAR(4)  
  lambda_BVAR <- 10 
  p_full<-4
  q_full<-0
}
if (VAR_type=="VAR")
{
# Marc's settings for VAR  
  p_full<-p
  q_full<-q
}

# Run the M-SSA wrapper, see tutorial 7.2
#   -The function computes M-SSA and M-MSE for each forecast horizon h in h_vec
if (recompute_results)
{
  final_mssa_indicator_obj<-compute_mssa_BIP_predictors_func(x_mat_wc,lambda_HP,L,date_to_fit_all,p_full,q_full,ht_mssa_vec,h_vec,f_excess,lag_vec,select_vec_multi,VAR_type,lambda_BVAR)
  save(final_mssa_indicator_obj,file=paste(path.result,"/final_mssa_indicator_obj",sep=""))
} else
{
  load(file=paste(path.result,"/final_mssa_indicator_obj",sep=""))
}
# Final M-SSA components
final_mssa_array<-final_mssa_indicator_obj$mssa_array

@@

colo<-c("blue","red")
mplot<-scale(mssa_final_vs_prem_wc)
mplot<-mplot[L:nrow(mplot),]
colnames(mplot)<-colnames(mssa_final_vs_prem_wc)
main_title<-paste("GDP component predictors: final and 2008-MSSA (h=",h,", shift=",shift,")",sep="")
plot(mplot[,1],main=main_title,axes=F,type="l",xlab="",ylab="",col=colo[1],ylim=c(min(na.exclude(mplot)),max(na.exclude(mplot))))
#mtext(colnames(mplot)[1],col=colo[1],line=-1)
for (jj in 1:ncol(mplot))
{
  lines(mplot[,jj],col=colo[jj],lwd=1,lty=1)
#  mtext(colnames(mplot)[jj],col=colo[jj],line=-jj)
}
abline(h=0)
axis(1,at=c(1,12*1:(nrow(mplot)/12)),labels=rownames(mplot)[c(1,12*1:(nrow(mplot)/12))])
axis(2)
legend("bottomleft", legend = colnames(mplot), col = colo, lty = 1, lwd = 1, bty = "n")
box()
invisible(dev.off())

@@

invisible(dev.off())

@@

\section{Hit Rate vs. False Alarm Rate}

%As shown in Wildi (2025), the SSA criterion's objective function—the target correlation $\rho(y,z,h)$—can be linked to the sign accuracy of the predictor $SA(y_t):=\text{P}\left(\text{sign}(z_{t+h}) = \text{sign}(y_t)\right)$, i.e., the probability that the signs of the predictor $y_t$ and the target $z_{t+h}$ agree:
%\begin{eqnarray}\label{sig_acc}
%SA(y_t)=0.5+\frac{\arcsin(\rho(y,z,h))}{\pi}.
%\end{eqnarray}
%Due to the strict monotonicity of the arcsin function, maximizing the target correlation also maximizes the sign accuracy of the predictor.\footnote{Equation \ref{sig_acc} holds exactly for Gaussian processes. Due to the central limit theorem, the relationship holds approximately for low-pass filtered non-Gaussian series, see Wildi (2025).} 
\begin{itemize}
\item Alternative to MSE, target correlation, holding-time: hit-rate, false alarm. 
\item We can link objective function (target correlation) with sign accuracy, see Wildi (2025). BUT: risk that reviewer asks for comparison with logit or whatever sign-forecast tool.
\item Context ROC-curve
\end{itemize}
For the ROC curves, we consider predictors of GDP based on both original (unfiltered) and filtered indicators. For the direct forecast, we use the regression
\[
GDP_{t+shift}=\beta_0^{shift}+\sum_{i=1}^5\beta_i^{shift}x_{ti}+\epsilon_{t}^{shift,direct}=\hat{GDP}_{t+shift}^{direct}+\epsilon_{t}^{shift,direct},
\]
where $x_{ti}$ designates the original indicators. 
For the filtered indicators, we consider similar regression equations:
\[GDP_{t+shift}=\beta_0^{shift,h,filt}+\sum_{i=1}^5\beta_i^{shift,h,filt}y_{ti}^{h,filt}+\epsilon_{t}^{h,shift,filt}=\hat{GDP}_{t+shift}^{filt}+\epsilon_{t}^{shift,filt},\]
where $y_{ti}^{h,filt}$ represents the $i$-th filtered indicator, based on either HP-C, M-MSE, or M-SSA. For simplicity of exposition, we present results based on the full sample, excluding the pandemic period. Fig. 
<<label=init,results=hide>>=

#########################################################################################################
# Apply ROC to direct forecast, direct HP forecast, M-MSE and M-SSA
# Data: 
#   -In-sample vs. out-of-sample
#   -Same sample (account for different NAs of various designs)
# Target: 
#   -GDP vs. HP-GDP
#   -forward-shift against h

# 1. Full sample without pandemic; all explanatories

# 1.1. Compute direct forecasts for h in h_vec
# Select all indicators
select_direct_indicator<-select_vec_multi
# Select data matrix: full sample without pandemic
data_roc<-x_mat_wc
direct_forecast_mat<-NULL
for (h in 0:max(h_vec))#h<-0
{
  # Shift GDP forward by publication lag+forecast horizon
  forward_shifted_GDP<-c(data_roc[(1+lag_vec[1]+h):nrow(data_roc),"BIP"],rep(NA,h+lag_vec[1]))
  # Regress selected indicators on forward-shifted GDP: start at t=L (same sample as HP-C)
  lm_obj<-lm(forward_shifted_GDP[L:length(forward_shifted_GDP)]~data_roc[L:nrow(data_roc),select_direct_indicator])
  summary(lm_obj)
  # Compute the predictor: one can rely on the generic R-function predict or compute the predictor manually
  direct_forecast<-lm_obj$coef[1]+data_roc[,select_direct_indicator]%*%lm_obj$coef[2:(length(select_direct_indicator)+1)]
  # Note that this is a full-sample predictor (no out-of-sample span)

  # We can now plot target and direct forecast: for h>2 the predictor comes close to a flat line centered at zero
  direct_forecast_mat<-cbind(direct_forecast_mat,direct_forecast)
}
colnames(direct_forecast_mat)<-paste("h=",h_vec,sep="")

#-----------
# 1.2. Direct HP forecast
direct_hp_forecast_mat<-forward_shifted_GDP_mat<-NULL
for (h in 0:max(h_vec))
{
  # Shift GDP forward by publication lag+forecast horizon
  forward_shifted_GDP<-c(data_roc[(1+lag_vec[1]+h):nrow(data_roc),"BIP"],rep(NA,h+lag_vec[1]))
  forward_shifted_GDP_mat<-cbind(forward_shifted_GDP_mat,forward_shifted_GDP)
  hp_mat<-NULL
  for (i in 1:ncol(data_roc))
   hp_mat<-cbind(hp_mat,filter(data_roc[,i],hp_c,side=1))
  colnames(hp_mat)<-select_vec_multi
  lm_obj<-lm(forward_shifted_GDP~hp_mat[,select_direct_indicator])
  # Compute the predictor: one can rely on the generic R-function predict or compute the predictor manually
  direct_hp_forecast_mat<-cbind(direct_hp_forecast_mat,lm_obj$coef[1]+hp_mat[,select_direct_indicator]%*%lm_obj$coef[2:(length(select_direct_indicator)+1)])
}
colnames(direct_hp_forecast_mat)<-colnames(forward_shifted_GDP_mat)<-paste("h=",h_vec,sep="")

#--------------
# 1.3. M-SSA and M-MSE
final_mssa_array<-final_mssa_indicator_obj$mssa_array
final_mmse_array<-final_mssa_indicator_obj$mmse_array

mssa_mat<-mmse_mat<-NULL
for (h in 0:max(h_vec))
{
  # Shift GDP forward by publication lag+forecast horizon
  forward_shifted_GDP<-c(data_roc[(1+lag_vec[1]+h):nrow(data_roc),"BIP"],rep(NA,h+lag_vec[1]))
  forward_shifted_GDP_mat<-cbind(forward_shifted_GDP_mat,forward_shifted_GDP)
  lm_obj<-lm(forward_shifted_GDP~t(final_mssa_array[select_direct_indicator,,h+1]))
# Compute the predictor: one can rely on the generic R-function predict or compute the predictor manually
  mssa_mat<-cbind(mssa_mat,lm_obj$coef[1]+t(final_mssa_array[select_direct_indicator,,h+1])%*%lm_obj$coef[2:(length(select_direct_indicator)+1)])
  lm_obj<-lm(forward_shifted_GDP~t(final_mmse_array[select_direct_indicator,,h+1]))
  mmse_mat<-cbind(mmse_mat,lm_obj$coef[1]+t(final_mmse_array[select_direct_indicator,,h+1])%*%lm_obj$coef[2:(length(select_direct_indicator)+1)])
}
colnames(mssa_mat)<-colnames(mmse_mat)<-paste("h=",h_vec,sep="")

#-------------
# 1.4. Targets: forward-shifted GDP and HP-GDP
forward_shifted_GDP_mat<-forward_shifted_GDP_mat
forward_shifted_HP_GDP_mat<-NULL
for (h in 0:max(h_vec))
{
  # Shift GDP forward by publication lag+forecast horizon
  forward_shifted_GDP<-c(data_roc[(1+lag_vec[1]+h):nrow(data_roc),"BIP"],rep(NA,h+lag_vec[1]))
  forward_shifted_HP_GDP_mat<-cbind(forward_shifted_HP_GDP_mat,filter(forward_shifted_GDP,hp_two,side=2))
}
colnames(forward_shifted_HP_GDP_mat)<-paste("shift",h_vec,sep="")

#-------------------------
# Apply ROC: generate curves and compute AUCs

# Combination of forward-shift (of target) and forecast horizon (underlying optimization)
shift_vec<-1:5
hh_vec<-1:6
AUC_array_GDP<-AUC_array_HP_GDP<-array(dim=c(length(shift_vec),length(hh_vec),4))
dimnames(AUC_array_GDP)<-dimnames(AUC_array_HP_GDP)<-list(paste("shift=",shift_vec,sep=""),paste("h=",hh_vec,sep=""),c("Direct forecast","Direct HP forecast","M-MSE","M-SSA"))
# Select shifts>k (otherwise too many plots in single window)
k<-2
# Show plot
showl<-T
# Labels: hit rate vs. false alarm rate
lbls = "Hit"

#----------------------
# Target GDP
select_target<-"GDP"

#file = "./Figures/ROC_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))

for (i in 1:length(shift_vec))#i<-1
{ 
  for (j in 1:length(hh_vec))#j<-1
  {
    shift<-shift_vec[i]
    h<-hh_vec[j]

# Select target: shifted GDP or HP-GDP
    if (select_target=="GDP")
    {
      target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
    } else

    {
      target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
    }
# Set up data matrix for ROC calculation    
    ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
    rownames(ROC_data)<-rownames(data_roc)
    colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
    ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
    showROC = ifelse(i>k&j>k,T,F)
    if (showROC&showl)
    {
      showLegend<-T
      showl<-F
    } else
    {
      showLegend<-F
    }
# Apply smoothing    
    smoothROC<-T

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
    AUC_array_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

#--------------------------
# Select  HP-GDP
select_target<-"HP-GDP"

#file = "./Figures/ROC_HP_GDP_many.pdf"
#pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(length(which(shift_vec>k)),length(which(hh_vec>k))))

for (i in 1:length(shift_vec))#i<-1
{ 
  for (j in 1:length(hh_vec))#j<-1
  {
    shift<-shift_vec[i]
    h<-hh_vec[j]

# Select target: shifted GDP or HP-GDP
    if (select_target=="GDP")
    {
      target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
    } else

    {
      target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
    }
# Set up data matrix for ROC calculation    
    ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
    rownames(ROC_data)<-rownames(data_roc)
    colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
    ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
    showROC = ifelse(i>k&j>k,T,F)
    if (showROC&showl)
    {
      showLegend<-T
      showl<-F
    } else
    {
      showLegend<-F
    }
# Apply smoothing    
    smoothROC<-T

    AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,", h=",h,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
    AUC_array_HP_GDP[i,j,]<-unlist(AUC)
  }
}
#invisible(dev.off())

#-----------------------------
# Shifts 1 and 4 only

select_target<-"GDP"

file = "./Figures/ROC_GDP_shift_1_4.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(1,2))

shift<-1
h<-1

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-T
AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
shift<-4
h<-shift

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-F
AUC<-ROCplots(ROC_data, showROC , main = paste("shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)

invisible(dev.off())

# Shifts 1 and 4 only

select_target<-"HP-GDP"

file = "./Figures/ROC_HP_GDP_shift_1_4.pdf"
pdf(file = paste(path.out,file,sep=""), paper = "special", width = 7, height = 7)
par(mfrow=c(1,2))

shift<-1
h<-1

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-T
AUC<-ROCplots(ROC_data, showROC , main = paste("HP-GDP: shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)
shift<-4
h<-shift

# Select target: shifted GDP or HP-GDP
if (select_target=="GDP")
{
  target<-as.integer(forward_shifted_GDP_mat[,shift+1]>0)
} else
{
  target<-as.integer(forward_shifted_HP_GDP_mat[,shift+1]>0)
}
# Set up data matrix for ROC calculation    
ROC_data<-cbind(target,direct_forecast_mat[,h+1],direct_hp_forecast_mat[,h+1],mmse_mat[,h+1],mssa_mat[,h+1])
rownames(ROC_data)<-rownames(data_roc)
colnames(ROC_data)<-c("Target","Direct forecast","Direct HP forecast","M-MSE","M-SSA")
ROC_data<-as.data.frame(na.exclude(ROC_data))
# Select plots that will be shown    
showLegend<-F
AUC<-ROCplots(ROC_data, showROC , main = paste("HP-GDP: shift=",shift,sep=""), lbls = lbls,
                  smoothROC , colours =NULL, lwd = 2,
                  showLegend , lg_cex = 1, lg_ncol = 1)

invisible(dev.off())

#----------------------------
# LOOK at AUCs

AUC_array_GDP["shift=1",,]
AUC_array_GDP["shift=2",,]
AUC_array_GDP["shift=3",,]
AUC_array_GDP["shift=4",,]
AUC_array_GDP["shift=5",,]

# Compute table with diagonal (shift=h)
auc_GDP_mat<-auc_HP_GDP_mat<-NULL
for (i in 1:length(shift_vec))
{
  auc_GDP_mat<-rbind(auc_GDP_mat,AUC_array_GDP[i,i,])
  auc_HP_GDP_mat<-rbind(auc_HP_GDP_mat,AUC_array_HP_GDP[i,i,])
}
rownames(auc_GDP_mat)<-rownames(auc_HP_GDP_mat)<-paste("Forecast horizon ",shift_vec,sep="")
auc_GDP_mat
auc_HP_GDP_mat

<<label=z_box_plot_pure_mba_2.pdf,echo=FALSE,results=tex>>=
file = "./Figures/ROC_GDP_shift_1_4.pdf"
cat("\\begin{figure}[H]")
cat("\\begin{center}")
cat("\\includegraphics[height=3in, width=5in]{", file, "}\n",sep = "")
cat("\\caption{Hit rate vs. false alarm rate of the predictors at small (one quarter ahead:left panel) and yearly forecast horizons (right panel).", sep = "")
cat("\\label{ROC_GDP_shift_1_4}}", sep = "")
cat("\\end{center}")
cat("\\end{figure}")
@

mat1<-round(auc_GDP_mat,3)
#latex(cor_vec, dec = 1, , caption = "Example of using latex to create table",
#center = "centering", file = "", floating = FALSE)
xtable(mat1, dec = 1,digits=rep(3,dim(mat1)[2]+1),
paste("AUCs of different predictors against GDP at various forecast horizons.."),
label=paste("p_val_wc5",sep=""),
center = "centering", file = "", floating = FALSE)
@
